---
title: "Predicting football matches"
output: html_document
---
TODO: add in team attributes to model data


```{r setup}
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
library(formattable)
library(lubridate)
library(foreach)
library(doParallel)
library(randomForest)
library(data.table)
library(tictoc)
library(caret)
```

##Initial look at data

```{r data}
dat<-src_sqlite("D:\\Downloads\\soccer\\database.sqlite",create = F)
src_tbls(dat)
```

##Getting match data

```{r data_matches,warning=FALSE}
matches<-tbl(dat,"Match") %>%  data.frame()
matches1<-filter(matches,is.na(home_player_X1)==F)
matches1$date<-ymd(as.POSIXct(matches1$date))

#calculating which team won 

matches1$result<-ifelse(matches1$home_team_goal>matches1$away_team_goal,"Home",                              ifelse(matches1$home_team_goal<matches1$away_team_goal,"Away","Draw"))
matches_player<-matches1 %>% 
  gather(key="Column",value="Value",-id) %>% 
  filter(grepl("player",Column)==T) %>% 
  filter(grepl("X|Y",Column)==F)
colnames(matches_player)[3]<-"player_api_id"

formattable(head(matches))
```

##Getting player data

This code gets the list of players in each match.

```{r players_matches_data}
players_in_matches<-filter(tbl(dat,"Player"),player_api_id%in%matches_player$player_api_id) %>%data.frame () %>% select(-id)

players_in_matches$player_api_id<-as.character(players_in_matches$player_api_id)
players_1<-left_join(matches_player,players_in_matches) %>% filter(is.na(player_api_id)==F)

players_1$match_date<-matches1[match(players_1$id,matches1$id),6]

formattable(head(players_1))
```

This section gets the players stats, at the time of their match.

```{r player_data}
player_attributes<-filter(tbl(dat,"Player_Attributes"),player_api_id%in%players_1$player_api_id) %>% data.frame()
player_attributes$date<-ymd(as.POSIXct(player_attributes$date))
player_attributes$player_api_id<-as.character(player_attributes$player_api_id)
player_attributes<-data.table(player_attributes)
setkey(x = player_attributes,player_api_id)

#transforming the data frame into a data table, to make is faster to filter the data
players_1<-data.table(players_1)
setkey(players_1,player_api_id)

formattable(head(players_1))

#This function matches the players attributes to the match data

match_player_and_attributes<-function(player,players = players_1,attributes = player_attributes){
  dat1<-players[which(players$player_api_id==player),]
  att_1<-attributes[which(attributes$player_api_id==player),] %>% 
    select(-id)
  dat2<-left_join(dat1,att_1)
  dat2$diff<-abs(dat2$match_date-dat2$date)
  dat2$paste<-paste(dat2$id,dat2$diff,sep="_")
  dat3<-dat2 %>% group_by(id) %>% 
    dplyr::summarise(diff=min(diff)) %>% 
    mutate("paste"=paste(id,diff,sep="_"))
  dat4<-filter(dat2,paste%in%dat3$paste) %>% select(-diff,-paste)
  return(dat4)
}

playersList<-unique(players_1$player_api_id)

#using a foreach loop to apply the function over all players
#creating parallel backend
registerDoParallel(makeCluster(2))
players_2<-foreach(a = 1:length(playersList),.combine = "rbind",.packages = c("data.table","dplyr"),.export = c("players_1","player_attributes"))%dopar%{
  match_player_and_attributes(playersList[a])
}


```

We can then calculate the average stats for each team

```{r av per team,warning=FALSE}
homeOrAway<-data.frame(str_split_fixed(string = players_2$Column,pattern = "_",2),stringsAsFactors = F)

players_2$homeOrAway<-homeOrAway$X1


players_grouped<-players_2 %>% 
  select(-player_api_id,-player_name,-player_fifa_api_id,-birthday,-Column,-match_date,-defensive_work_rate,-attacking_work_rate,-date,-preferred_foot) %>% 
  gather(key="Metric",value="Value",-id,-homeOrAway) %>% 
  mutate("Value"=as.numeric(Value)) %>% 
  group_by(id,homeOrAway,Metric) %>% 
  summarise("Mean"=mean(Value,na.rm = T),"Median"=median(Value,na.rm = T),"Max"=max(Value,na.rm = T))

players_grouped_2<-players_grouped %>% 
  ungroup() %>% 
  gather(key="Column",value="Value",-id,-homeOrAway,-Metric) %>% 
  mutate("Col"=paste(Metric,Column,homeOrAway,sep="_")) %>% 
  select(-Column,-Metric,-homeOrAway) %>% 
  filter(is.na(Value)==F) %>% 
  filter(is.infinite(Value)==F) %>% 
  spread(key="Col",value="Value",fill=0)
```

We can then join this up with the match data

```{r join}
matches2<-select(matches1,id,home_team_api_id,away_team_api_id,result)
joinedData<-left_join(matches2,players_grouped_2,by="id")

```

##Predicting results

```{r ramndomForest}
#getting two thirds of the data
rows<-sample(nrow(joinedData),nrow(joinedData)*2/3)
trainData<-joinedData[rows,]
testData<-joinedData[-rows,]

trainDataMinus<-select(trainData,-result,-id)
resultsTrain<-select(trainData,result) %>% 
  mutate("result"=as.factor(result))
testDataMins<-select(testData,-result,-id)
testResult<-select(testData,result) %>% 
  mutate("result"=as.factor(result))

set.seed(123)

model<-randomForest(x = trainDataMinus,xtest =testDataMins ,ytest = testResult$result,y = resultsTrain$result)

testResult$Predicted<-model$test$predicted

testResult$Correct<-ifelse(testResult$result==testResult$Predicted,1,0)

sum(testResult$Correct)/nrow(testResult)
```

```{r summarising results}
results<-testResult %>% 
  group_by(Predicted,result) %>% 
  summarise(Count=n())
```

```{r team attributes}
team_att<-tbl(dat,"Team_Attributes") %>%
  data.frame() %>% 
  gather(key="Column",value="Value",-id,-team_fifa_api_id,-team_api_id,-date) %>% 
  select(-id,-team_fifa_api_id) 
team_att$date<-ymd(as.POSIXct(team_att$date))

matches3<-select(matches1,id,date,home_team_api_id,away_team_api_id,result)

matchTeamAttr<-function(matchID,matchData,team_att=team_att){
  dat1<-filter(matchData,id==matchID)
  homeID<-dat1[1,3]
  awayID<-dat1[1,4]
  homeStat<-filter(team_att,team_api_id==homeID) %>% 
    filter(date<dat1$date)%>% 
      mutate("id"=matchID)
    
  if(nrow(homeStat)>0){
    homeStat<-filter(homeStat,date==max(homeStat$date)) 
  }
    awayStat<-filter(team_att,team_api_id==awayID) %>% 
    filter(date<dat1$date)%>% 
      mutate("id"=matchID)
  if(nrow(awayStat)>0){
    awayStat<-filter(awayStat,date==max(awayStat$date)) 
  }
    dat2<-full_join(homeStat,awayStat,by=names(homeStat))
    return(dat2)
}
#aa<-matchTeamAttr(15014,matches3,team_att)

team_stats<-foreach(a = 1:nrow(matches3),.combine = "rbind",.errorhandling = "remove")%do%{
  matchTeamAttr(matches3[a,1],matches3,team_att)
}

match_team_stats<-matchData %>% 
  gather(key="Col",value="team_api_id",-id,-date,-result) 
match_team_stats_1<-team_stats %>% 
  select(-date) %>% 
  left_join(match_team_stats)
matchTeam<-data.frame(str_split_fixed(match_team_stats_1$Col,"_",2),stringsAsFactors = F)

match_team_stats_1$Col<-matchTeam$X1
match_team_stats_1$Column<-paste(match_team_stats_1$Col,match_team_stats_1$Column,sep="_")

#joining up with player data
match_team_player_stats<-match_team_stats_1 %>% 
  select(-Col,-date,-team_api_id) %>% 
  spread(key="Column",value="Value",fill=0) %>% 
  left_join(players_grouped_2)

#forcing character strings to be vectors

for(a in 1:ncol(match_team_player_stats)){
  if(is.na(sum(as.numeric(match_team_player_stats[,a])))){
    match_team_player_stats[,a]<-as.factor(match_team_player_stats[,a])
  }else{
    match_team_player_stats[,a]<-as.numeric(match_team_player_stats[,a])
  }
}
```

We can then check this with a model, and see if out predictions have improved

```{r model with team data}
rows<-sample(nrow(match_team_player_stats),nrow(match_team_player_stats)*2/3)

trainData1<-match_team_player_stats[rows,]
testData1<-match_team_player_stats[-rows,]

trainData1M<-select(trainData1,-result,-id)
testData1M<-select(testData1,-result,-id)

set.seed(100)
model1<-randomForest(x = trainData1M,xtest =testData1M ,ytest = testData1$result,y = trainData1$result)
```

Testing results

```{r model 1 test}

testResult1<-data.frame("Actual"=testData1$result,"Predicted"=model1$test$predicted,stringsAsFactors = F)

testResult1$Correct<-ifelse(testResult1$Actual==testResult1$Predicted,1,0)

sum(testResult1$Correct)

sum(testResult1$Correct)/nrow(testResult1)
```